Variety in the Implementation of Nonlinear Least Squares Program Codes

### 2022-05-27 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
Minor edits to get clean output

---
title: "Variety in the Implementation of Nonlinear Least Squares Program Codes"
author: 
   - John C Nash, retired professor, University of Ottawa
date: "16/02/2021 revised 20/01/2022"
output: 
    pdf_document:
        keep_tex: false
        toc: true
bibliography: ../BibSupport/ImproveNLS.bib
---

@@

There are many ways to structure a Gauss-Newton style nonlinear least squares
program code. In organizing and documenting the nearly half-century of programs
in the Nashlib collection associated with @jncnm79 and with similar packages
for the R Project for Statistical Computing, the author realized that an overview
of this variety could be instructive for numerical software designers.

@@

Given a function that can be written as a sum of squared residuals

where $x$ is a set of real valued parameters, we wish to find the minimum of $f(x)$
with respect to x. There may be additional conditions that add to the complexity of
the problem such as

@@

@@

An overview of this method is provided by Wikipedia: 
\url{https://en.wikipedia.org/wiki/Gauss%E2%80%93Newton_algorithm}
The method was described by C.F.Gauss in 1809 in *Theoria motus corporum 
coelestium in sectionibus conicis solem ambientum*.

In calculus we learn that a stationary point (local maximum, minimum
or saddle point) of a function $f$ occurs where its gradient (or first
derivative) is zero. In multiple dimensions, this is the same as
having the gradient $g$ stationary. The so-called **Newton method**
uses the second derivatives in the Hessian matrix $H$ defined as

@@

and we iterate until there is no change in $x$, assuming that the method does indeed
converge for the problem at hand. Newton (and his sometimes associated worker Raphson) 
used the algorithm we attribute
to them in a very special case and in a manner unrecognizable as that described
above.  (See \url{https://en.wikipedia.org/wiki/Newton%27s_method}).

For our present situation, the $f(x)$ can be written as a sum of squares as above. It is worth
repeating the vector form of this

$$  f(x) = r^t r $$

because it is less fussy to write and read. This is clear when we expand the evaluation
of the Hessian of this sum of squares function.

@@

Arguing that the residuals should be "small", Gauss proposed that the 
Newton equations could be approximated by ignoring
the second term (which have elements of the residual vector times second derivatives 
of the residual).  This gives the **Gauss-Newton**
equations (cancelling a factor 2)

@@

<!-- ?? Should we explain more. Possibly incorporate into code? -->

@@

can be thought of as yielding a step $\delta$ that merges the Gauss-Newton and 
steepest-descents direction. This approach was actually first suggested by 
@Levenberg1944, but it is my opinion that Levenberg 
likely never exercised them in practice, while Marquardt actually had a well-used
program. Before computers were commonly
available, many papers were written with computational
ideas that were given at most a cursory trial. (In 1984, I had the pleasure of showing
Marquardt my own BASIC program running on a Radio Shack TRS80 Model 100 and he  
commented on how the computations were converging to a solution as we watched
the iterations proceed.)

Note that we can define a matrix $K$ which is $J$ augmented below by a suitable diagonal
matrix that is the square root of $\lambda$ times the unit matrix $1_n$. Then the
Levenberg-Marquardt equations are equivalent to

@@

Many practitioners dislike using IF statements (comparisons) of floating point
numbers. Tolerances are often suggested to define "small", but this invites users
to make decisions that lead to poor outcomes and which ignore the precision and
other characteristics of the arithmetic at hand.
There are, of course, well-written codes that address the issues of what "small" means,
but such code is generally quite involved. The R function `all.equal()` presents
a well-considered approach. However,
I have never seen a case where the $OFFSET$ approach caused difficulties, and it
is trivial to code. When $x$ is tiny, we compare $OFFSET$ to 
itself and declare equality. When $x$ is very large, $OFFSET$ does not change it and
we compare $x$ with a value that is identical to it.

@@

<!-- ?? Check the statements about how things are done and redo tests -->
Note that we form $J^t J$ in `nlsr::nlfb()` to provide the $D$ matrix for 
the scaled stabilization, 
which is quite a lot of computational work. The same program, however, 
uses a QR approach to solving the linearized least squares sub-problem.
To avoid the formation of $D$, an experimental package
`nlsralt` was been built to add functions `nlxbx()` and `nlfbx()` that
use ONLY the fixed element (i.e., unit matrix) stabilization. A
preliminary timing (using @microbench13) with the Hobbs unscaled problem 
(@cnm79, Example 12.2), presented later, shows
a slight advantage for `nlsralt` in execution timing, but we caution
that this is but a single example.

@@

The main attraction of the Gauss-Newton method is that the Jacobian only needs first
derivatives, which are generally much, much easier to compute than the full second
derivatives of the Hessian. However, there have been suggested
Newton-like approaches, possibly for special types of models where some short cut
to the Hessian is available. I regard these approaches as part of the
computational infrastructure of the special cases, as it is usually 
difficult to apply them outside of their context.

Two such examples are the R packages `maxLik` (@maxLik2011) and `trust` (@p-trust).

<!-- expand ??, include some BHHH stuff -->

@@

We have a versions of the Nashlib Algorithm 23 in BASIC, Fortran, Pascal, and R, 
with Python pending. There may be dialects of these programming languages also,
giving rise to other variations.

@@

The `nls()` function has an argument `subset` and a test included in the 
`Croucher-example1.R` file of this project showed this works as expected. 
However, nowhere have we found a published example of the use of the 
`subset` argument with `nls()`. Worse, it clearly must be applied to
many of the functions used by `nls()` but makes no explicit appearance. Thus it
is VERY difficult to ensure subsetting is correctly applied when we wish to modify
the code, for instance, to provide for newer features especially more reliable
solvers.

Note that the `subset` ARGUMENT to `nls()` is not the same as the `subset()` function
that is usable to restrict the scope of an R object.

OUR HYPOTHESIS: `nls()` uses `subset` to automatically adjust the data to which all
the relevant functions are applied. We note that `minpack.lm::nlsLM` also has a `subset` argument,
but package `nlsr` does not at time of writing (2021-7-12), but will use the **weights** mechanism
of the next subsection of this document at some future release.

@@

In order to provide a sane and well-understood mechanism for subsetting, there
are alternatives for implementation.

- **via weights**: We can weight each "row" that we do NOT wish included as 0.
  This does require us to compute the number of observations as the count of
  non-zero weights e.g, `sum(weights > 0). Incidentally, this underlines the 
  need to ensure weights are non-negative. ??now in nlsj, need to add to nlsr code.??

- **via preprocessing**: We could prepare a function to pre-process the problem and
  create a data collection that is appropriately subset. This may offend against
  the spirit of R/S, but it has the merit that we can easily reassure ourselves
  that we have worked with the correct information, correctly used in our methods.
  ?? Should add this as an option.

PROPOSAL: use the "weights" approach if `nlsj` is called with `subset` active, 
else recommend the preprocessing approach. The preprocessor could also do a lot
of checking, which we might leave out of `nlsj()` in some cases. As of 2021-7-12,
the weights mechanism is working in a trial `nlsj()` and should be transferrable
to package `nlsr` fairly easily.

>>>>

---
title: "Variety in the Implementation of Nonlinear Least Squares Program Codes"
author: 
   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
date: "16/02/2021 revised 27/05/2022"
output: 
    pdf_document:
        keep_tex: false
        toc: true
bibliography: ../BibSupport/ImproveNLS.bib
---

@@

There are many ways to structure a Gauss-Newton style nonlinear least squares
program code. In organizing and documenting the nearly half-century of programs
in the Nashlib collection associated with @jncnm79 and with similar packages
for the R Project for Statistical Computing, the authors realized that an overview
of this variety could be instructive for numerical software designers.

@@

Numerical analysts will generally want to solve a **nonlinear least squares** problem,
that is, minimize an objective function with respect to a set of $n$ parameters, $x$, where the 
objective  $f()$ can be written as a sum of $m$ squared elements.

$$  f(x, Y) = \sum_{i=1}^m {r_i(x,Y)^2} $$

where $Y$ represents the ensemble of **exogenous** data that is not changed by alteration
of the parameters $x$. The functions $r_i()$ need not have the same form. An example is
the well know Rosenbrock banana shaped valley, which has $n = n = 2$ and

$$ r_1(x) = 10*(x_2 - x_1 ^ 2) $$

$$ r_2(x) = (1 - x_1) $$

There is no exogenous data here.

There may be additional conditions that add to the complexity of
the problem such as

@@

Many other researchers will want to solve a **nonlinear regression** problem, where
the $r_i()$ are all of the same form. The (unscaled) Hobbs weed infestation problem (@jncnm79, page 120) uses

$$ r_i(x) = x_1/(1+x_2 * exp(-x_3 * i)) - y_i$$

where the exogenous data is a single vector of numbers $y$. In nonlinear regression, we are mostly
interested in "fitting a model to data", and the "minimizing a sum of squares" is the approach.

While the term "nonlinear" devolves from functions where parameters appear to degree 1, a more
useful classification is according to whether a fixed number of computations can estimate the
parameters. That is, problems can be considered "nonlinear" if we cannot compute the best parameter values
in a pre-defined fixed number of computations and must use an iterative approach.

In what follows, we will primarily take the numerical analyst's viewpoint. That is, we will be looking
mostly at the calculations for the minimization of the sum of squares. However, the purpose of these
calculations can dictate much of the structure of the software in which such computations appear.

@@

An overview of the method is provided by Wikipedia: *https://en.wikipedia.org/wiki/Gauss%E2%80%93Newton_algorithm*.

The method was described by C.F.Gauss in 1809 in *Theoria motus corporum 
coelestium in sectionibus conicis solem ambientum*.

In calculus we learn that a stationary point (local maximum, minimum
or saddle point) of a function $f$ occurs where its gradient (or first
derivative) is zero. In multiple dimensions, this is the same as
having the gradient $g$ stationary. The so-called Newton method
uses the second derivatives in the Hessian matrix $H$ defined as

@@

and we iterate until there is no change in $x$, assuming that the method does indeed
converge for the problem at hand. Newton (and his sometimes associated worker Raphson) 
used the algorithm we attribute
to them in a very special case and in a manner unrecognizable as that described
above.

There are some adjustments when our function $f(x)$ can be written as a sum of squares

The final vector product form is the one I favour because it is the least fussy to write and read.
In particular, when we try to evaluate the Hessian of this sum of squares function, we see that it is

@@

Arguing that the residuals should be "small", Gauss proposed that the 
Newton equations could be approximated by ignoring
the second term (which has elements of the residual vector times second derivatives 
of the residual).  This gives the **Gauss-Newton**
equations (cancelling a factor 2)

@@

@@

can be thought of as yielding a step $\delta$ that merges the Gauss-Newton and 
steepest-descents direction. This approach was actually first suggested by 
@Levenberg1944, but it is my opinion that Levenberg 
likely never exercised them in practice, while Marquardt actually had a well-used
program. Before computers were commonly
available, many papers were written with computational
ideas that were given at most a cursory trial. (In 1984, JN had the pleasure of showing
Marquardt his own BASIC version of this program running on a Radio Shack TRS80 Model 100 and 
Marquardt commented on how the computations were converging to a solution as we watched
the iterations proceed.)

Note that we can define a matrix $K$ which is $J$ augmented below by a suitable diagonal
matrix that is the square root of $\lambda$ times the unit matrix $1_n$. Then the
Levenberg-Marquardt equations are equivalent to solving the linear least squares problem

@@

Many practitioners dislike using IF statements (comparisons) of floating point
numbers. Tolerances are often suggested to define "small", but this invites users
to make decisions that lead to poor outcomes and which ignore the precision and
other characteristics of the arithmetic at hand.
There are, of course, well-written codes that address the issues of what "small" means,
but such code is generally quite involved. The R function `all.equal()` presents
a well-considered approach. However,
We have never seen a case where the $OFFSET$ approach caused difficulties, and it
is trivial to code. When $x$ is tiny, we compare $OFFSET$ to 
itself and declare equality. When $x$ is very large, $OFFSET$ does not change it and
we compare $x$ with a value that is identical to it.

@@

Note that we form $J^t J$ in `nlsr::nlfb()` to provide the $D$ matrix for 
the scaled stabilization, 
which is quite a lot of computational work. The same program, however, 
uses a QR approach to solving the linearized least squares sub-problem.
To avoid the formation of $D$, an experimental package
`nlsralt` has been built to add functions `nlxbx()` and `nlfbx()` that
use ONLY the fixed element (i.e., unit matrix) stabilization. A very
preliminary timing ([2021-6-26]) using the Hobbs unscaled problem shows
a slight improvement for the simplified version in timing with equivalent
solution. More extensive testing is, we feel, warranted to show if this
approach offers a useful gain in performance while preserving robustness
in finding the parameter estimates.

@@

Newton (and his sometime associated worker Raphson) used the algorithm we attribute
to them applied to a very special case and in a manner unrecognizable as that described
above.  (See *https://en.wikipedia.org/wiki/Newton%27s_method*).

The main attraction of the Gauss-Newton method is that the Jacobian only needs first
derivatives, which are generally much, much easier to compute than the full second
derivatives of the Hessian. However, there have been suggestioned
Newton-like approaches, possibly for special types of models where some short cut
to the Hessian is available. These may or may not make sense in the context of
a given user or group of users. I regard these approaches as part of the
computational infrastructure of the special cases rather than as general nonlinear
least squares methods. The `maxLik` package (@maxLik2011) is a prime example.

@@

We have a versions of the Nashlib Algorithm 23 in BASIC, Fortran, Pascal, Matlab/Octave and R, 
with Python under consideration. There may be dialects of these programming languages,
giving rise to other variations. Features of these computing environments can greatly
ease or hinder the building of well-structured codes. See
*git@github.com:pcolsen/Nash-Compact-Numerical-Methods.git*.

@@

The `nls()` function has an argument `subset`.  Here is an example of its use.

```{r croucher}
# Croucher-example1.R -- https://walkingrandomly.com/?p=5254

xdata = c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9)
ydata = c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001)
Cform <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata) # formula
Cstart<-list(p1=1,p2=0.2) # starting parameters
Cdata<-data.frame(xdata, ydata) # data frame of the data
Csubset<-3:8 # index of rows to use from data frame

# fit using all the data
fitfull<-nls(Cform, data=Cdata, start=Cstart)
fitfull
# fit using subset data
fitsub<-nls(Cform, data=Cdata, subset=Csubset, start=Cstart)
fitsub
```

A more extensive version of this with explicit extraction of the data 
showed this works as expected. 
However, nowhere have we found a published example of the use of the 
`subset` argument with `nls()`. Worse, it appears to be applied to
many of the functions used by `nls()` but makes no explicit appearance. Thus it
is VERY difficult to ensure subseting is correctly applied when we wish to modify
the code, for instance, to provide for more reliable solvers.

Note that the `subset` ARGUMENT to `nls()` is not the same as the `subset()` function
that is usable to restrict the scope of an R object. The potential confusion that a
common name implies is an additional concern.

@@

In order to provide a sane and well-understood mechanism for subseting, there
are alternatives for implementation.

- **via weights**: We can weight each "row" that we do NOT wish included as 0.
  This does require us to compute the number of observations as the count of
  non-zero weights e.g, `sum(weights > 0). Incidentally, this underlines the 
  need to ensure weights are non-negative.

- **via preprocessing**: We could prepare a function to pre-process the problem and
  create a dataframe that is appropriately subset. This may offend against
  the spirit of R/S, but it has the merit that we can easily reassure ourselves
  that we have worked with the correct information.

  <!-- ?? Should add this as an option. -->

<!-- PROPOSAL: use the "weights" approach if `nlsj` is called with `subset` active,  -->
<!-- else recommend the preprocessing approach. The preprocessor could also do a lot -->
<!-- of checking, which we might leave out of `nlsj()` in some cases. As of 2021-7-12, -->
<!-- the weights mechanism is working in a trial `nlsj()` and should be transferrable -->
<!-- to package `nlsr` fairly easily. -->
